---
title: "final project"
format: html
editor: visual
---

## Summary


```{r}
install.packages("pdftools")
install.packages("textdata") 
install.packages("topicmodels")
install.packages("reshape2")

library(pdftools)
library(tidyverse)
library(cowplot)
library(tesseract)
library(dplyr)
library(stringr)
library(tidytext)
library(topicmodels)
library(reshape2)
```

## Step 1: Read in the PDFs

This analysis uses Letters and Agreements of Postsecondary Disability Discrimination cases in the state of Indiana. Download the PDFs here: <https://ocrcas.ed.gov/ocr-search?f%5B0%5D=it%3APost%20Secondary&f%5B1%5D=ocr_statutes%3A528&f%5B2%5D=state_fullname%3A653>.

Let's read them in using the `pdftools` package and store them in a single data frame.

```{r}
# Set base path
base_path <- "/Users/jennamareewong/Library/CloudStorage/Box-Box/Courses/SOC 384J : Computational Social Science/Final Project/Data/"

# Get all PDF files in the directory
pdf_files <- list.files(base_path, pattern = "\\.pdf$", full.names = TRUE)

# Read all PDFs into a named list
pdf_texts <- lapply(pdf_files, pdf_text)

# Name each list element by the filename (without path and extension)
names(pdf_texts) <- tools::file_path_sans_ext(basename(pdf_files))
```

Create lists of the letters and agreements.

```{r}
agreements <- pdf_texts[grepl("_Agreement", names(pdf_texts))]
letters    <- pdf_texts[grepl("_Letter",    names(pdf_texts))]
```

Create a data frame.

```{r}
# Create the data frame from pdf_texts
pdf_df <- map_dfr(names(pdf_texts), function(doc_name) {
  tibble(
    document  = doc_name,
    page      = 1:length(pdf_texts[[doc_name]]),
    text      = pdf_texts[[doc_name]]
  )
})

# Add doc_type column
pdf_df <- pdf_df |>
  mutate(doc_type = case_when(
    grepl("_Agreement", document) ~ "Agreement",
    grepl("_Letter",    document) ~ "Letter"
  ))
```

## Step 2: Tokenization

Converting text into a **tidy** format: one token (word) per row. The `unnest_tokens()` function does this for us.

```{r}
## tokenize: split text into individual words
agreements_tokens <- pdf_df %>%
  filter(doc_type == "Agreement") %>%
  unnest_tokens(word, text)

letters_tokens <- pdf_df %>%
  filter(doc_type == "Letter") %>%
  unnest_tokens(word, text)

## look at the result
agreements_tokens
letters_tokens
```

## Step 3: Remove stop words

Stop words are common words like "the", "and", "is" that don't carry much meaning. The `tidytext` package includes a built-in list of stop words.

```{r}
## look at the stop words list
data(stop_words)
head(stop_words, 10)
```

We use an `anti_join()` to remove stop words — this keeps only the words that are *not* in the stop words list.

```{r}
## see top words to help determine stop words
agreements_tokens %>%
  count(word, sort = TRUE) %>%
  head(50)

letters_tokens %>%
  count(word, sort = TRUE) %>%
  head(50)

## create custom stop words
custom_stopwords <- tibble(
  word = c("department", "education", "office", "university", "office", "ocr", "agreement", "college", "xxxxx", "xxxxxx", "including")
)

## remove stop words
agreements_clean <- agreements_tokens %>%
  anti_join(stop_words, by = "word") %>%
  anti_join(custom_stopwords, by = "word")

letters_clean <- letters_tokens %>%
  anti_join(stop_words, by = "word") %>%
  anti_join(custom_stopwords, by = "word")
```

## Step 4: Word frequencies

Now let's count the most common words in each set of reviews.

```{r}
## count words by review set
ag_word_counts <- agreements_clean %>%
  count(document, word, sort = TRUE)

let_word_counts <- letters_clean %>%
  count(document, word, sort = TRUE)
```

Let's visualize the top 15 words for letters and for agreements:

```{r}
## plot top 15 words per case
let_word_counts %>%
  group_by(document) %>%
  slice_max(n, n = 10) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, n, document)) %>%
  ggplot(aes(x = n, y = word)) +
  geom_col(show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~document, scales = "free_y") +
  scale_y_reordered() +
  theme_cowplot() +
  labs(x = "Count", y = NULL, title = "Most common words by discrimination case letter")

ag_word_counts %>%
  group_by(document) %>%
  slice_max(n, n = 10) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, n, document)) %>%
  ggplot(aes(x = n, y = word)) +
  geom_col(show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~document, scales = "free_y") +
  scale_y_reordered() +
  theme_cowplot() +
  labs(x = "Count", y = NULL, title = "Most common words by discrimination case agreement")
```

## Step 5: Dictionary methods

A dictionary method is simple: we define a list of words that belong to a category, and then we count how often those words appear. This is one of the most straightforward approaches to text analysis.

Let's create a custom dictionary to classify reviewer language into categories relevant to peer review.

```{r}
## define a custom peer review dictionary
review_dictionary <- tibble(
  word = c(
    ## type of disability
    "visual", "overfitting", "validation", "validate", "bias", "error",
    "errors", "residuals", "uncertainty", "robust", "robustness",
    ## positive evaluation
    "impressive", "excellent", "important", "interesting", "valuable", "strong",
    "clear", "nice", "enjoyed", "wonderful", "helpful", "strength", "strengths",
    ## negative evaluation
    "lacking", "disappointing", "flawed", "problematic", "poor", "poorly",
    "weak", "missing", "unclear", "insufficient", "outdated", "concern", "concerns",
    ## suggestions
    "suggest", "suggestion", "improve", "improvement", "consider",
    "recommend", "request", "include", "adding", "clarify"
  ),
  category = c(
    rep("Methodology", 11),
    rep("Positive", 13),
    rep("Negative", 13),
    rep("Suggestion", 10)
  )
)

review_dictionary
```

Now let's apply the dictionary by joining it with our tokenized reviews:

```{r}
## apply dictionary using inner_join
dictionary_results <- reviews_clean %>%
  inner_join(review_dictionary, by = "word")

## count by review set and category
dictionary_summary <- dictionary_results %>%
  count(review_set, category)

dictionary_summary
```

```{r}
## visualize dictionary results
dictionary_summary %>%
  ggplot(aes(x = category, y = n, fill = review_set)) +
  geom_col(position = "dodge", alpha = 0.8) +
  theme_cowplot() +
  scale_fill_viridis_d() + 
  labs(x = "Category", y = "Count", fill = "Review Set",
       title = "Dictionary-based classification of reviewer language") +
  theme(legend.position = "bottom")
```

We can also compute the share of dictionary words in each category per review set — this accounts for the fact that some reviews are longer than others.

```{r}
## compute proportions within each review set
dictionary_proportions <- dictionary_results %>%
  count(review_set, category) %>%
  group_by(review_set) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

dictionary_proportions %>%
  ggplot(aes(x = category, y = proportion, fill = review_set)) +
  geom_col(position = "dodge", alpha = 0.8) +
  theme_cowplot() +
    scale_fill_viridis_d() + 
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Category", y = "Share of dictionary words", fill = "Review Set",
       title = "Proportion of reviewer language by category") +
  theme(legend.position = "bottom")
```

```{r}
data(stop_words)

custom_stopwords <- tibble(
  word = c("department", "education", "office", "complaint", "recipient", "resolution", "agreement", "university", "office", "ocr")
)

tidy_corpus <- tidy_corpus %>%
  anti_join(stop_words, by = "word") %>% 
  anti_join(custom_stopwords, by = "word")
```

Count words per document.

```{r}
word_counts <- tidy_corpus %>%
  count(file, word, sort = TRUE)
```

Create document-term matrix.

```{r}
dtm <- word_counts %>%
  cast_dtm(document = file, term = word, value = n)
```

## Step 6: Sentiment analysis

Sentiment analysis uses pre-built dictionaries (lexicons) to classify words as positive or negative. The `tidytext` package gives us access to several sentiment lexicons.

### AFINN lexicon

The AFINN lexicon scores words on a scale from -5 (most negative) to +5 (most positive).

```{r}
## look at the AFINN lexicon
afinn <- get_sentiments("afinn")
head(afinn, 10)
```

```{r}
# Join with sentiment scores
afinn_scores_let <- letters_clean %>%
  inner_join(afinn, by = "word")

afinn_scores_ag <- agreements_clean %>%
  inner_join(afinn, by = "word")

# Compute average sentiment by document
afinn_summary_let <- afinn_scores_let%>%
  group_by(document) %>%
  summarize(
    mean_sentiment = mean(value),
    sd_sentiment = sd(value),
    n_words = n(),
    se = sd_sentiment / sqrt(n_words)
  )

afinn_summary_ag <- afinn_scores_ag%>%
  group_by(document) %>%
  summarize(
    mean_sentiment = mean(value),
    sd_sentiment = sd(value),
    n_words = n(),
    se = sd_sentiment / sqrt(n_words)
  )

afinn_summary_let
afinn_summary_ag
```

Plot the average sentiments with confidence intervals.

```{r}
# Letters
afinn_summary_let %>%
  ggplot(aes(x = document, y = mean_sentiment,
             ymin = mean_sentiment - 1.96 * se,
             ymax = mean_sentiment + 1.96 * se)) +
  geom_pointrange(shape = 21, fill = "white", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  theme_cowplot() +
  coord_flip() +
  labs(x = "Document", y = "Mean AFINN Sentiment Score",
       title = "Average sentiment by letter")

# Agreements
afinn_summary_ag %>%
  ggplot(aes(x = document, y = mean_sentiment,
             ymin = mean_sentiment - 1.96 * se,
             ymax = mean_sentiment + 1.96 * se)) +
  geom_pointrange(shape = 21, fill = "white", size = 1) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  theme_cowplot() +
  coord_flip() +
  labs(x = "Document", y = "Mean AFINN Sentiment Score",
       title = "Average sentiment by agreement")

# Merge into one figure
afinn_combined <- bind_rows(
  afinn_summary_let %>% mutate(doc_type = "Letter"),
  afinn_summary_ag %>% mutate(doc_type = "Agreement")
) %>%
  # Extract institution name by removing "_Letter" or "_Agreement" suffix
  mutate(institution = gsub("_Letter.*|_Agreement.*", "", document))

# Plot the new figure
afinn_combined %>%
  ggplot(aes(x = mean_sentiment, y = institution, color = doc_type,
             xmin = mean_sentiment - 1.96 * se,
             xmax = mean_sentiment + 1.96 * se)) +
  geom_pointrange(shape = 21, fill = "white", size = 1,
                  position = position_dodge(width = 0.5)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  theme_cowplot() +
  labs(y = "Institution", x = "Mean AFINN Sentiment Score",
       color = "Document Type",
       title = "Sentiment by institution: Letters vs Agreements")

```

### Bing lexicon

The Bing lexicon classifies words as simply "positive" or "negative."

```{r}
## look at bing lexicon
bing <- get_sentiments("bing")
head(bing, 10)

bing <- bing %>% filter(!word %in% c("inequality", "inequalities"))
```

```{r}

# --- LETTERS ---

letters_clean <- letters_clean %>%
  mutate(institution = gsub("_Letter.*", "", document))

agreements_clean <- agreements_clean %>%
  mutate(institution = gsub("_Agreement.*", "", document))

bing_scores_let <- letters_clean %>%
  inner_join(bing, by = "word")

bing_summary_let <- bing_scores_let %>%
  count(institution, sentiment)

bing_summary_let %>%
  ggplot(aes(x = institution, y = n, fill = sentiment)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("negative" = "darkred", "positive" = "steelblue")) +
  theme_cowplot() +
  coord_flip() +
  labs(x = "Institution", y = "Word Count", fill = "Sentiment",
       title = "Positive vs. negative words by letter (Bing lexicon)") +
  theme(legend.position = "bottom")

# --- AGREEMENTS ---
bing_scores_ag <- agreements_clean %>%
  inner_join(bing, by = "word")

bing_summary_ag <- bing_scores_ag %>%
  count(institution, sentiment)

bing_summary_ag %>%
  ggplot(aes(x = institution, y = n, fill = sentiment)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("negative" = "darkred", "positive" = "steelblue")) +
  theme_cowplot() +
  coord_flip() +
  labs(x = "Institution", y = "Word Count", fill = "Sentiment",
       title = "Positive vs. negative words by agreement (Bing lexicon)") +
  theme(legend.position = "bottom")

# --- TOP SENTIMENT WORDS ACROSS ALL DOCUMENTS ---
bing_scores_all <- bind_rows(bing_scores_let, bing_scores_ag)

bing_scores_all %>%
  count(word, sentiment, sort = TRUE) %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>%
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(x = n, y = word, fill = sentiment)) +
  geom_col(show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~sentiment, scales = "free_y") +
  scale_fill_manual(values = c("negative" = "indianred", "positive" = "steelblue")) +
  theme_cowplot() +
  labs(x = "Count", y = NULL, title = "Top sentiment words across all documents")
```

```{r}
## visualize
bing_summary %>%
  ggplot(aes(x = review_set, y = n, fill = sentiment)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("negative" = "darkred", "positive" = "steelblue")) +
  theme_cowplot() +
  labs(x = "Review Set", y = "Word Count", fill = "Sentiment",
       title = "Positive vs. negative words by review set (Bing lexicon)") + 
  theme(legend.position = "bottom")
```

Let's look at which specific words are driving the sentiment:

```{r}
## top positive and negative words across all reviews
bing_word_counts <- bing_scores %>%
  count(word, sentiment, sort = TRUE) %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>%
  ungroup()

bing_word_counts %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(x = n, y = word, fill = sentiment)) +
  geom_col(show.legend = FALSE, alpha = 0.8) +
  facet_wrap(~sentiment, scales = "free_y") +
  scale_fill_manual(values = c("negative" = "indianred", "positive" = "steelblue")) +
  theme_cowplot() +
  labs(x = "Count", y = NULL, title = "Top sentiment words across all reviews")
```

**Understanding check:** Look at the top "negative" words. Are all of them actually negative in the context of peer review? What does this tell us about the limitations of sentiment analysis?

### NRC lexicon

The NRC lexicon classifies words into 8 emotions (anger, anticipation, disgust, fear, joy, sadness, surprise, trust) plus positive/negative.

```{r}
## get nrc lexicon
nrc <- get_sentiments("nrc")

## join with tokens
nrc_scores <- reviews_clean %>%
  inner_join(nrc, by = "word")

## count emotions by review set (exclude positive/negative)
nrc_summary <- nrc_scores %>%
  filter(!sentiment %in% c("positive", "negative")) %>%
  count(review_set, sentiment)

## visualize
nrc_summary %>%
  ggplot(aes(x = sentiment, y = n, fill = review_set)) +
  geom_col(position = "dodge", alpha = 0.8) +
  theme_cowplot() +
  labs(x = "Emotion", y = "Count", fill = "Review Set",
       title = "NRC emotion classification by review set") +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0),
        legend.position = "bottom")
```

## Latent Dirichlet Allocation

Create the data
```{r}
dtm_ag <- ag_word_counts %>%
  cast_dtm(document = document, term = word, value = n)

dtm_let <- let_word_counts %>%
  cast_dtm(document = document, term = word, value = n)

```

Pull topics from LDA
```{r}
lda_model_ag <- LDA(dtm_ag, k = 5, control = list(seed = 1234))
lda_model_let <- LDA(dtm_let, k = 5, control = list(seed = 1234))

```

Analyze topics
```{r}
topics_let <- tidy(lda_model_let, matrix = "beta")
topics_ag <- tidy(lda_model_ag, matrix = "beta")

doc_topics_let <- tidy(lda_model_let, matrix = "gamma")

dominant_topics_let <- doc_topics_let %>%
  group_by(document) %>%
  slice_max(gamma, n = 1)

top_terms_let <- topics_let %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  ungroup()

top_terms_let %>%
  ggplot(aes(x = reorder_within(term, beta, topic), y = beta)) +
  geom_col() +
  facet_wrap(~ topic, scales = "free") +
  coord_flip() +
  scale_x_reordered()

top_terms_let

doc_topics_ag <- tidy(lda_model_ag, matrix = "gamma")

dominant_topics_ag <- doc_topics_ag %>%
  group_by(document) %>%
  slice_max(gamma, n = 1)

top_terms_ag <- topics_ag %>%
  group_by(topic) %>%
  slice_max(beta, n = 10) %>%
  ungroup()

top_terms_ag %>%
  ggplot(aes(x = reorder_within(term, beta, topic), y = beta)) +
  geom_col() +
  facet_wrap(~ topic, scales = "free") +
  coord_flip() +
  scale_x_reordered()

top_terms_ag

```


